Associahedra for finite type cluster algebras \\ and minimal relations between $\b{g}$-vectors

### 2023-02-21 thirdPaper/draft/typeConeAssociahedra.tex

% operators
\DeclareMathOperator{\conv}{conv} % convex hull
\DeclareMathOperator{\vect}{vect} % linear span
\DeclareMathOperator{\cone}{cone} % cone hull
\DeclareMathOperator{\rank}{rank} % rank

@@

\smallskip
\begin{itemize}
\item \textbf{any initial seed}, regardless of whether it is acyclic or not (see \cref{exm:typeConeCA,fig:clusterFans,fig:labelFacetDefiningInequalititiesCA1,fig:labelFacetDefiningInequalititiesCA2} for examples in types~$A_3$ and~$C_3$~cyclic). In contrast, the proof of~\cite{BazierMatteDouvilleMousavandThomasYildirim} only treats acyclic initial seeds (although H.~Thomas announced in a lecture given at RIMS in June 2019 that this restriction will disappear in a future version of~\cite{BazierMatteDouvilleMousavandThomasYildirim}).

\smallskip
\item \textbf{all finite type cluster algebras}, regardless of whether it is simply-laced or not (see \cref{exm:typeConeCA} for an example in type~$C_3$). In contrast, even with the acyclicity restriction, the result of~\cite{BazierMatteDouvilleMousavandThomasYildirim} is first proved in simply-laced cases ($A$, $D$ and~$E$) using representation theory of quivers. Extension to the non simply-laced cases is then argued by a folding argument. Although not presented in detail, this argument is subtle and technical.

@@

\begin{proposition}
\label{prop:characterizationPolytopalFan}
Let~$\Fan$ be an essential complete simplicial fan in~$\R^n$. Then the following are equivalent for any height vector~$\b{h}\in \R^N$:
\begin{enumerate}
\item The fan~$\Fan$ is the normal fan of the polytope~$P_\b{h} \eqdef \set{\b{x} \in \R^n}{\b{G}\b{x} \le \b{h}}$.
\item For any two adjacent maximal cones~$\R_{\ge0}\rays$ and~$\R_{\ge0}\rays'$ of~$\Fan$ with~$\rays \ssm \{\ray\} = \rays' \ssm \{\ray'\}$,
\[
\alpha \, \b{h}_{\ray} + \alpha' \, \b{h}_{\ray'} + \sum_{\ray[s] \in \rays \cap \rays'} \beta_{\ray[s]} \, \b{h}_{\ray[s]} > 0,
\]
where
\[
\alpha \, \ray + \alpha' \, \ray' + \sum_{\ray[s] \in \rays \cap \rays'} \beta_{\ray[s]} \, \ray[s] = 0
\]
is the unique (up to rescaling) linear dependence with~$\alpha, \alpha' > 0$ between the rays of~$\rays \cup \rays'$.
\end{enumerate}
\end{proposition}

@@

\begin{example}
\label{exm:typeCone}
\begin{figure}[b]
	\capstart
	\centerline{{\includegraphics[scale=1]{typeConeFan}} \quad {\includegraphics[scale=1]{typeConePolytope}} \quad \raisebox{.5cm}{\includegraphics[scale=.45]{typeCone}}}
	\caption{A $2$-dimensional fan~$\Fan$ with five rays $1, \dots, 5$ and five maximal cones~$a, \dots, e$ (left), its polytopal realization corresponding to the height vector~$(\nicefrac{1}{2}, \nicefrac{3}{4}, 1, 1, \nicefrac{5}{4})$ (middle), and the intersection of $\b{K}\typeCone(\Fan)$ with a hyperplane to get a $2$-dimensional slice (right).}
	\label{fig:typeCone}
\end{figure}
%
Consider the $2$-dimensional fan~$\Fan$ depicted in \cref{fig:typeCone} (left) (we will encounter this fan again in \cref{exm:typeConeCoarsenings,exm:typeConeMinkowskiSum,exm:typeConeAsso}).
It has five rays labeled~$1, \dots, 5$ and five maximal cones labeled~$a,\dots, e$.
For the matrices $\b{G}$ and $\b{K}$, we consider
\[
%\transpose{\b{G}} = \begin{bmatrix} 1 & 0 & -1 & -1 & 0 \\ 0 & 1 & 1 & 0 & -1 \end{bmatrix}
\b{G} = \begin{bmatrix} 1 & 0 \\ 0 & 1 \\ -1 & 1 \\ -1 & 0 \\ 0 & -1 \end{bmatrix}
\qquad\text{and}\qquad
\b{K} = \begin{bmatrix} 1 & 0 & 0 & 1 & 0 \\ 0 & 1 & 0 & 0 & 1 \\ 1 & 0 & 1 & 0 & 1 \end{bmatrix}.
\]
The type cone~$\typeCone(\Fan)$ lies in~$\R^5$, but has a $2$-dimensional lineality space.
The five pairs of adjacent maximal cones of~$\Fan$ give rise to following five defining inequalities for~$\typeCone(\Fan)$:
\begin{gather*}
H^>_{ab}: \; \b{h}_1 + \b{h}_3 - \b{h}_2 > 0
\qquad
H^>_{bc}: \; \b{h}_2 + \b{h}_4 - \b{h}_3 > 0
\qquad
H^>_{cd}: \; \b{h}_3 + \b{h}_5 - \b{h}_4 > 0
\\
H^>_{de}: \; \b{h}_1 + \b{h}_4 > 0
\qquad
H^>_{ae}: \; \b{h}_2 + \b{h}_5 > 0,
\end{gather*}
where $H^>_{ij}$ denotes the halfspace defined by the inequality corresponding to the two adjacent maximal cones~$i$ and~$j$.
Note that the inequalities~$H^>_{ae}$ and~$H^>_{de}$ are redundant.
For example, the height vector~$(\nicefrac{1}{2}, \nicefrac{3}{4}, \nicefrac{5}{4}, 1, \nicefrac{5}{4})$ belongs to the type cone~$\typeCone(\Fan)$, and the corresponding polytope is represented in \cref{fig:typeCone} (middle).
To represent this type cone, it suffices to represent~$\b{K}\typeCone(\Fan)$ which is an essential simplicial cone in~$\R^3$, given as the intersection of the following five open halfspaces:
\begin{gather*}
\b{K}H^>_{ab}: \; {x} - {y} +2 {z} > 0
\qquad
\b{K}H^>_{bc}: \; {x} + {y} - {z} > 0
\qquad
\b{K}H^>_{cd}: \; -{x} + {y} +2 {z} > 0
\\
\b{K}H^>_{de}: \; 2{x} + {z} > 0
\qquad
\b{K}H^>_{ae}: \; 2{y} + {z} > 0.
\end{gather*}
One can further reduce the dimension by intersecting with a transversal hyperplane to get a $2$-dimensional representation.
This is the red triangle depicted in \cref{fig:typeCone} (right).
\end{example}

@@

\pagebreak
We now focus on the normal fan of~$\Asso[n]$.
Since~$\Asso[n]$ lives in a hyperplane of~$\R^{n+1}$, its normal fan has a one-dimensional lineality.
Let~$\HH \eqdef \bigset{\b{x} \in \R^{n+1}}{\sum_{\ell \in [n+1]} \b{x}_\ell = 0}$ and ${\pi : \R^{n+1} \to \HH}$ denote the orthogonal projection. We denote by~$\gvectorFan[n]$ the intersection of the normal fan of~$\Asso[n]$ with~$\HH$, which is an essential complete simplicial fan.

>>>>

% operators
\DeclareMathOperator{\conv}{conv} % convex hull
\DeclareMathOperator{\vect}{vect} % linear span
\DeclareMathOperator{\cone}{cone} % cone hull
\DeclareMathOperator{\rank}{rank} % rank
\DeclareMathOperator{\oprays}{rays} % rays

@@

\smallskip
\begin{itemize}
\item \textbf{any initial seed}, regardless of whether it is acyclic or not (see \cref{exm:typeConeCA,fig:clusterFans,fig:labelFacetDefiningInequalititiesCA1,fig:labelFacetDefiningInequalititiesCA2} for examples in types~$A_3$ and~$C_3$~cyclic). In contrast, the proof of~\cite{BazierMatteDouvilleMousavandThomasYildirim} only treats acyclic initial seeds {\color{red} although the approach has since been extended to any seed in~\cite{BazierMatte-thesis} (as previously announced by H.~Thomas in a lecture given at RIMS in June 2019).}

\smallskip
\item \textbf{all finite type cluster algebras}, regardless of whether it is simply-laced or not (see \cref{exm:typeConeCA} for an example in type~$C_3$). In contrast, even with the acyclicity restriction, the result of~\cite{BazierMatteDouvilleMousavandThomasYildirim} is first proved in simply-laced cases ($A$, $D$ and~$E$) using representation theory of quivers. {\color{red} Extension to the non simply-laced cases can then be argued by a folding argument. This subtle and technical argument is presented in~\cite{ArkaniHamedHeLam}.}

@@

\begin{proposition}
\label{prop:characterizationPolytopalFan}
Let~$\Fan$ be an essential complete simplicial fan in~$\R^n$ with $N$ rays. Then the following are equivalent for any height vector~$\b{h}=\left(\b{h}_{\ray}\right)_{\ray \in \oprays(\Fan)}\in \R^N$:
\begin{enumerate}
\item The fan~$\Fan$ is the normal fan of the polytope~$P_\b{h} \eqdef \set{\b{x} \in \R^n}{\b{G}\b{x} \le \b{h}}$.
\item For any two adjacent maximal cones~$\R_{\ge0}\rays$ and~$\R_{\ge0}\rays'$ of~$\Fan$ with~$\rays \ssm \{\ray\} = \rays' \ssm \{\ray'\}$,
\[
\alpha \, \b{h}_{\ray} + \alpha' \, \b{h}_{\ray'} + \sum_{\ray[s] \in \rays \cap \rays'} \beta_{\ray[s]} \, \b{h}_{\ray[s]} > 0,
\]
where
\[
\alpha \, \ray + \alpha' \, \ray' + \sum_{\ray[s] \in \rays \cap \rays'} \beta_{\ray[s]} \, \ray[s] = 0
\]
is the unique (up to rescaling) linear dependence with~$\alpha, \alpha' > 0$ between the rays of~$\rays \cup \rays'$.
\end{enumerate}
\end{proposition}

@@

\begin{example}
\label{exm:typeCone}
\begin{figure}[b]
	\capstart
	\centerline{{\includegraphics[scale=1]{typeConeFan}} \quad {\includegraphics[scale=1]{typeConePolytope}} \quad \raisebox{.5cm}{\includegraphics[scale=.45]{typeCone}}}
	\caption{A $2$-dimensional fan~$\Fan$ with five rays $1, \dots, 5$ and five maximal cones~$a, \dots, e$ (left), its polytopal realization corresponding to the height vector~$(\nicefrac{1}{2}, \nicefrac{3}{4}, 1, 1, \nicefrac{5}{4})$ (middle), and the intersection of $\b{K}\typeCone(\Fan)$ with a hyperplane to get a $2$-dimensional slice (right).}
	\label{fig:typeCone}
\end{figure}
%
Consider the $2$-dimensional fan~$\Fan$ depicted in \cref{fig:typeCone} (left) (we will encounter this fan again in \cref{exm:typeConeCoarsenings,exm:typeConeMinkowskiSum,exm:typeConeAsso}).
It has five rays labeled~$1, \dots, 5$ and five maximal cones labeled~$a,\dots, e$.
For the matrices $\b{G}$ and $\b{K}$, we consider
\[
%\transpose{\b{G}} = \begin{bmatrix} 1 & 0 & -1 & -1 & 0 \\ 0 & 1 & 1 & 0 & -1 \end{bmatrix}
\b{G} = \begin{bmatrix} 1 & 0 \\ 0 & 1 \\ -1 & 1 \\ -1 & 0 \\ 0 & -1 \end{bmatrix}
\qquad\text{and}\qquad
\b{K} = \begin{bmatrix} 1 & 0 & 0 & 1 & 0 \\ 0 & 1 & 0 & 0 & 1 \\ 1 & 0 & 1 & 0 & 1 \end{bmatrix}.
\]
The type cone~$\typeCone(\Fan)$ lies in~$\R^5$, but has a $2$-dimensional lineality space.
The five pairs of adjacent maximal cones of~$\Fan$ give rise to following five defining inequalities for~$\typeCone(\Fan)$:
\begin{gather*}
H^>_{ab}: \; \b{h}_1 + \b{h}_3 - \b{h}_2 > 0
\qquad
H^>_{bc}: \; \b{h}_2 + \b{h}_4 - \b{h}_3 > 0
\qquad
H^>_{cd}: \; \b{h}_3 + \b{h}_5 - \b{h}_4 > 0
\\
H^>_{de}: \; \b{h}_1 + \b{h}_4 > 0
\qquad
H^>_{ae}: \; \b{h}_2 + \b{h}_5 > 0,
\end{gather*}
where $H^>_{ij}$ denotes the halfspace defined by the inequality corresponding to the two adjacent maximal cones~$i$ and~$j$.
Note that the inequalities~$H^>_{ae}$ and~$H^>_{de}$ are redundant.
For example, the height vector~$(\nicefrac{1}{2}, \nicefrac{3}{4}, \nicefrac{5}{4}, 1, \nicefrac{5}{4})$ belongs to the type cone~$\typeCone(\Fan)$, and the corresponding polytope is represented in \cref{fig:typeCone} (middle).
To represent this type cone, it suffices to represent~$\b{K}\typeCone(\Fan)$, which is an essential simplicial cone in~$\R^3$, given as the intersection of the following five open halfspaces:
\begin{gather*}
\b{K}H^>_{ab}: \; {x} - {y} +2 {z} > 0
\qquad
\b{K}H^>_{bc}: \; {x} + {y} - {z} > 0
\qquad
\b{K}H^>_{cd}: \; -{x} + {y} +2 {z} > 0
\\
\b{K}H^>_{de}: \; 2{x} + {z} > 0
\qquad
\b{K}H^>_{ae}: \; 2{y} + {z} > 0.
\end{gather*}
One can further reduce the dimension by intersecting with a transversal hyperplane to get a $2$-dimensional representation. Any affine hyperplane crossing all the rays of~$\b{K}\typeCone(\Fan)$ would work, here we take the hyperplane $x+y+z=1$ and we get the red triangle depicted in \cref{fig:typeCone} (right).
\end{example}

@@

\pagebreak
We now focus on the normal fan of~$\Asso[n]$.
Since~$\Asso[n]$ lives in a hyperplane of~$\R^{n+1}$, its normal fan has a one-dimensional lineality (spanned by the all-ones vector).
To factor out the lineality, let~$\HH \eqdef \bigset{\b{x} \in \R^{n+1}}{\sum_{\ell \in [n+1]} \b{x}_\ell = 0}$ and ${\pi : \R^{n+1} \to \HH}$ denote the orthogonal projection.
We denote by~$\gvectorFan[n]$ the intersection of the normal fan of~$\Asso[n]$ with~$\HH$. Note that~$\gvectorFan[n]$ is an $n$-dimensional essential complete simplicial fan consisting of the projections under~$\pi$ of the normal cones of $\Asso[n]$. With the choice of a basis, we can identify $\HH$ with $\R^n$ and depict~$\Asso[n]$ as a full-dimensional polytope with normal fan~$\gvectorFan[n]$.
